Convergence rate of MAP estimates \\ for the exponential family

### 2021-05-19 main.tex
Add empirical illustration of bias variance decomposition

\begin{example}
	\begin{theorem}[MLE tight upper bound]
	The MLE of $\cN(0,\mu_*)$ is $\hat \mu_n^\text{MLE} = \inv{n} \sum_i X_i^2 $.
	Its expected suboptimality is infinite when $n\leq 2$, and otherwise upper-bounded as
	\begin{align}
		 \expect{\bregmanconj( \mu_*; \hat \mu_n^\text{MLE}) }
			\leq \inv{2n} +\frac{2}{n(n-2)} \; .
			\label{eq:tight_bounds}
	\end{align}
\end{theorem}
Note that this upper bound is asymptotically tight.
 There is also a closed form for the multivariate generalization, thanks to the \href{https://en.wikipedia.org/wiki/Inverse-Wishart_distribution}{inverse Wishart distribution} and the \href{https://en.wikipedia.org/wiki/Wishart_distribution#Log-expectation}{expectation of the log-determinant of a Wishart}. The expected value is infinite whenever $n \leq d+1$ where $d$ is the dimension. We do not perform these calculations for now as we want to focus on the simplest possible examples.

As for the MAP, we did not manage to get an asymptotically tight upper bound. Nevertheless, using inequality~\eqref{eq:log_bound}, we did get an interesting upper bound. 
To start, let us recall that the MAP of $\cN(0,\mu_*)$ is $\hat \mu_n = \frac{n_0 \mu_0 + \sum_i X_i^2}{n_0 + n}$.
Its expectation is simply $\mu_n= \frac{n_0 \mu_0 + n \mu^*}{n_0 + n}$.
We now present a lemma on its expected inverse.
\begin{lemma}[Expected MAP natural parameter]
	For any $n\geq 2$, the expectation of the natural parameter of the MAP of $\cN(0,\mu_*)$ is bounded as
	\begin{align}
		\frac{\mu^*}{\mu_n}
		\leq \expect{\frac{\mu^*}{\MAPm}} 
		= \expect{\frac{\MAPt}{\natp^*}} 
		\leq \frac{1}{\frac{\mu_n}{\mu^*} - \frac{2}{n+n_0}} \; .
	\end{align}
\end{lemma}
Not that the lower bound is a simple consequence of the convexity of the inverse function. Using this lemma, along with~\eqref{eq:log_bound}, we get the following result.
\begin{theorem}[MAP Upper Bound]
 Introducing $a=n_0 \frac{\mu_0}{\mu^*}$, the expected suboptimality of the MAP of $\cN(0,\mu^*)$ with prior hyper-parameters $(n_0,\mu_0)$ is 
\begin{equation}
	\expect{\bregmanconj( \mu_*; \hat \mu_n^\text{MAP})}
	\leq \begin{cases}
		\bregmanconj( \mu_*; \mu_0) \ \text{ if } \ n=0, \\
		\bregmanconj( \mu_*; \mu_0) + \inv{a} \ \text{if}\ n=1,\\
		\frac{1}{a+n-2} + \frac{\half(1 - \frac{\mu_0}{\mu^*})^2}{(\frac{\mu_0}{\mu^*}+\frac{n-2}{n_0})(1 + \frac{n}{n_0} )} \ \text{if}\  n\geq 2 \ \text{and}\ a=n_0 \frac{\mu_0}{\mu^*} >0
	\end{cases}
\end{equation}
\end{theorem}

@@

@@

\subsection{MLE Loose Upper Bound}
If we apply \eqref{eq:log_bound} to the MLE, we get the upper bound
\begin{align}
	\expect{\bregmanconj(\mu^*; \MAPm^\text{MLE})}
	&\leq  \half \left( \expect{\frac{n}{\chi^2(n)}} + \expect{\frac{\chi^2(n)}{n}} - 2 \right ) \\
	&= \half \left( \frac{n}{n-2} + 1 - 2 \right)  
	= \frac{1}{n-2}
\end{align}
which is exactly  an additive term $1/2n$ larger than \eqref{eq:tight_bounds}. This highlights the factor $2$ difference between the log-bound and an exact computation of the expectation of the log term, and a tight upper bound on the digamma function.

@@

\section{Bias-Variance Decomposition}
Using Jensen inequality and inserting appropriate quantities, we obtain the following analog of the bias-variance decomposition:
\newcommand{\bias}[1]{\textcolor{Red}{\underbrace{\textcolor{black}{#1}}_{\text{bias}}}}
\newcommand{\variance}[1]{\textcolor{Green}{\underbrace{\textcolor{black}{#1}}_{\text{variance}}}}
\begin{align}
	& \expect{\bregmanconj(\mu^* ; \hat \mu_n)} 
	= \conj(\mu^* ) 
	- \underbrace{\expect{\conj(\hat \mu_n)}}_{\leq \conj(\mu_n) \text{ (Jensen)} } 
	\pm \langle \underbrace{\nabla \conj( \mu_n)}_{=:\natp_n} ; \mu^* - \mu_n \rangle
	+ \E[\langle \hat \natp_n ; \hat \mu_n   \pm \mu_n - \mu^*  \rangle ] \\
	& \leq \conj(\mu^* ) 
	- \conj(\mu_n) 
	- \langle  \nabla \conj( \mu_n) ; \mu^* - \mu_n \rangle
	+ \langle \E[\hat \natp_n] - \natp_n ;  \mu_n - \mu^* \rangle
	+ \E[\langle \hat \natp_n ; \hat \mu_n - \mu_n \rangle ]
	\nonumber\\
%	& \quad \quad \quad 
%	- \inv{n+n_0} \smallsum_i \E[\langle \hat \natp_n ; \mu^* - T(X_i)  \rangle]\\
	& \leq \bias{\bregmanconj(\mu^*; \mu_n)}
	+ \frac{n_0}{n+n_0} \langle  \bias{\mu_0 - \mu_*} ; \variance{\E[\hat \natp_n] - \natp_n} \rangle 
	+ \variance{\Cov(\hat \natp_n ; \hat \mu_n ) }
\end{align}
where $\hat \mu_n = \frac{n_0 \mu_0 + \sum_i T(X_i)}{n_0 + n}$ is the MAP estimate of the mean parameter, $\hat \natp_n = \nabla \conj(\hat \mu_n)$ is the corresponding natural parameter, 
$\mu_n = \expect{\hat \mu_n} = \frac{n_0 \mu_0 + n \mu^*}{n_0 + n}$ is the expected MAP estimate,
and $\theta_n = \nabla \conj(\mu_n)$ is the natural parameter of this expectation.
Note that unless $\conj$ is $\ell^2$, or we are in a degenerate case with $\hat \mu_n$ a Dirac, $\theta_n \neq \expect{\hat \theta_n}$.

@@

@@

\begin{example}
	We can also get a closed form for the posterior of a centered gaussian variance 
	\begin{align}
		\expect[\pi(\natp|\cD)]{\bregman(\natp ; \natp_*)} = \bregman(\hat \natp_n ; \natp_*) + \frac{2}{n} \frac{\mu_*}{\hat \mu_n} + \log\frac{n}{2} - \psi(1 + \frac{n}{2})
	\end{align}
	using the mean of the \href{https://en.wikipedia.org/wiki/Inverse-gamma_distribution}{inverse Gamma distribution} and the \href{https://en.wikipedia.org/wiki/Gamma_distribution#Logarithmic_expectation_and_variance}{logarithmic expectation} of a Gamma. The using results from the MLE, we get the exact formula
	\begin{align}
		\expect[\cD\sim\natp_*]{\expect[\pi(\natp|\cD)]{\bregman(\natp ; \natp_*)}}
		= \frac{2}{n} + \frac{2}{n(n-2)} + \psi(\half[n]) - \psi(1+ \half[n])
	\end{align}
	and we bound it to 
	\begin{align}
		\inv{2n} + \frac{4}{n(n-2)}
		\leq \expect[\cD\sim\natp_*]{\expect[\pi(\natp|\cD)]{\bregman(\natp ; \natp_*)}}
		\leq 
		\inv{n} + \frac{4}{n(n-2)}
	\end{align}
	which is about twice the MLE bounds~\eqref{eq:tight_bounds}, exactly like as we observed for the gaussian mean.
\end{example}

>>>>

\begin{example}
	\begin{theorem}[MLE tight upper bound]
	The MLE of $\cN(0,\mu_*)$ is $\hat \mu_n^\text{MLE} = \inv{n} \sum_i X_i^2 $.
	Its expected suboptimality is infinite when $n\leq 2$, and otherwise upper-bounded as
	\begin{align}
		 \expect{\bregmanconj( \mu_*; \hat \mu_n^\text{MLE}) }
			\leq \inv{2n} +\frac{2}{n(n-2)} \; .
			\label{eq:MLE_rate}
	\end{align}
\end{theorem}
Note that this upper bound is asymptotically tight.
 There is also a closed form for the multivariate generalization, thanks to the \href{https://en.wikipedia.org/wiki/Inverse-Wishart_distribution}{inverse Wishart distribution} and the \href{https://en.wikipedia.org/wiki/Wishart_distribution#Log-expectation}{expectation of the log-determinant of a Wishart}. The expected value is infinite whenever $n \leq d+1$ where $d$ is the dimension. We do not perform these calculations for now as we want to focus on the simplest possible examples.

As for the MAP, we did not manage to get an asymptotically tight upper bound. Nevertheless, using inequality~\eqref{eq:log_bound}, we did get an interesting upper bound. 
To start, let us recall that the MAP of $\cN(0,\mu_*)$ is $\hat \mu_n = \frac{n_0 \mu_0 + \sum_i X_i^2}{n_0 + n}$.
Its expectation is simply $\mu_n= \frac{n_0 \mu_0 + n \mu^*}{n_0 + n}$.
We now present a lemma on its expected inverse.
\begin{lemma}[Expected MAP natural parameter]
	For any $n\geq 2$, the expectation of the natural parameter of the MAP of $\cN(0,\mu_*)$ is bounded as
	\begin{align}
		\frac{\mu^*}{\mu_n}
		\leq \expect{\frac{\mu^*}{\MAPm}} 
		= \expect{\frac{\MAPt}{\natp^*}} 
		\leq \frac{1}{\frac{\mu_n}{\mu^*} - \frac{2}{n+n_0}} \; .
	\end{align}
\end{lemma}
Not that the lower bound is a simple consequence of the convexity of the inverse function. Using this lemma, along withthe los upper bound on the logarithm~\eqref{eq:log_bound}, we get a convergence rate for the MAP. We relate this rate convergence rate with the rate of MLE in Figure~\ref{fig:MAP_rate}.
\begin{theorem}[MAP Upper Bound]
 Introducing $a=n_0 \frac{\mu_0}{\mu^*}$, the expected suboptimality of the MAP of $\cN(0,\mu^*)$ with prior hyper-parameters $(n_0,\mu_0)$ is 
\begin{equation}
	\expect{\bregmanconj( \mu_*; \hat \mu_n^\text{MAP})}
	\leq \begin{cases}
		\bregmanconj( \mu_*; \mu_0) \ \text{ if } \ n=0, \\
		\bregmanconj( \mu_*; \mu_0) + \inv{a} \ \text{if}\ n=1,\\
		\frac{1}{a+n-2} + \frac{\half(1 - \frac{\mu_0}{\mu^*})^2}{(\frac{\mu_0}{\mu^*}+\frac{n-2}{n_0})(1 + \frac{n}{n_0} )} \ \text{if}\  n\geq 2 \ \text{and}\ a=n_0 \frac{\mu_0}{\mu^*} >0
	\end{cases}
	\label{eq:MAP_rate}
\end{equation}
\end{theorem}
Using the same bound on the log~\eqref{eq:log_bound}, we can also get a loose bound on the MLE, which is useful for comparison purposes.
\begin{equation}
	\expect{\bregmanconj( \mu_*; \hat \mu_n^\text{MLE}) }
	\leq \inv{n-2} \; .
	\label{eq:MLE_loose_rate}
\end{equation}

@@

\begin{figure}[ht]
	\centering
	\includegraphics[width=.7\textwidth]{figs/MAP_rates/MAP_rate_n0=5.pdf}
	\caption{
	Convergence rate of MAP~\eqref{eq:MAP_rate} 
	\label{fig:MAP_rate} against the ratio between initialization and optimum $ \frac{\mu_0}{\mu^*}$. 
	We also report the tight convergence rate of MLE~\eqref{eq:MLE_rate} (dash-dotted horizontal lines), and the loose upper bound~\eqref{eq:MLE_loose_rate} (horizontal lines). 
	We report these curves for various values of $n$, and $n_0=5$. 
	We observe that the rate of MAP is significantly better (lower) than the rate of MLE, only when we guessed $\mu_0$ right, within an order of magnitude from $\mu^*$, and for a small number of sample ($n<10$). 
	This observation hints towards the optimality of MLE for large $n$. 
	It makes me feel very interested in convergence rates of MAP in the overparametrized regime $n<\dim(T(X))$.
	}
\end{figure}

@@

\subsection{MLE Loose Upper Bound}
If we apply \eqref{eq:log_bound} to the MLE, we get the upper bound
\begin{align}
	\expect{\bregmanconj(\mu^*; \MAPm^\text{MLE})}
	&\leq  \half \left( \expect{\frac{n}{\chi^2(n)}} + \expect{\frac{\chi^2(n)}{n}} - 2 \right ) \\
	&= \half \left( \frac{n}{n-2} + 1 - 2 \right)  
	= \frac{1}{n-2}
\end{align}
which is exactly  an additive term $1/2n$ larger than \eqref{eq:MLE_rate}. This highlights the factor $2$ difference between the log-bound and an exact computation of the expectation of the log term, and a tight upper bound on the digamma function.

@@

\section{Bias-Variance Decomposition}
\paragraph{Covariance.}
Using Jensen inequality and inserting appropriate quantities, we obtain the following analog of the bias-variance decomposition:
\newcommand{\bias}[1]{\textcolor{Red}{\underbrace{\textcolor{black}{#1}}_{\text{bias}}}}
\newcommand{\variance}[1]{\textcolor{Green}{\underbrace{\textcolor{black}{#1}}_{\text{variance}}}}
\begin{align}
	& \expect{\bregmanconj(\mu^* ; \hat \mu_n)} 
	= \conj(\mu^* ) 
	- \underbrace{\expect{\conj(\hat \mu_n)}}_{\leq \conj(\mu_n) \text{ (Jensen)} } 
	\pm \langle \underbrace{\nabla \conj( \mu_n)}_{=:\natp_n} ; \mu^* - \mu_n \rangle
	+ \E[\langle \hat \natp_n ; \hat \mu_n   \pm \mu_n - \mu^*  \rangle ] \nonumber \\
	& \leq \conj(\mu^* ) 
	- \conj(\mu_n) 
	- \langle  \nabla \conj( \mu_n) ; \mu^* - \mu_n \rangle
	+ \langle \E[\hat \natp_n] - \natp_n ;  \mu_n - \mu^* \rangle
	+ \E[\langle \hat \natp_n ; \hat \mu_n - \mu_n \rangle ]
	\nonumber \\
	& \leq \bias{\bregmanconj(\mu^*; \mu_n)}
	+ \frac{n_0}{n+n_0} \langle  \bias{\mu_0 - \mu_*} ; \variance{\E[\hat \natp_n] - \natp_n} \rangle 
	+ \variance{\Cov(\hat \natp_n ; \hat \mu_n ) }
\end{align}
where $\hat \mu_n = \frac{n_0 \mu_0 + \sum_i T(X_i)}{n_0 + n}$ is the MAP estimate of the mean parameter, $\hat \natp_n = \nabla \conj(\hat \mu_n)$ is the corresponding natural parameter, 
$\mu_n = \expect{\hat \mu_n} = \frac{n_0 \mu_0 + n \mu^*}{n_0 + n}$ is the expected MAP estimate,
and $\theta_n = \nabla \conj(\mu_n)$ is the natural parameter of this expectation.
Note that unless $\conj$ is $\ell^2$, or we are in a degenerate case with $\hat \mu_n$ a Dirac, $\theta_n \neq \expect{\hat \theta_n}$.

@@

\begin{figure}[ht]
	\centering
	\includegraphics[width=\textwidth]{figs/numerical_schema.pdf}
	\caption{A numerical illustration of the different characters featured in the bias-variance decomposition, for a 1D Gaussian $\cN(\mu, \sigma^2)$.}
	\label{fig:bias-variance-numerical}
\end{figure}

\paragraph{No Jensen.}
In fact this previous decomposition is highly suboptimal. 
We do not need to use Jensen inequality.
\begin{align}
	\expect{\bregmanconj(\mu^* ; \hat \mu_n)} 
	&= \conj(\mu^* ) 
	\pm \conj(\mu_n)
	-  \expect{\conj(\hat \mu_n)}
	+ \E[\langle \hat \natp_n ; \hat \mu_n   \pm \mu_n - \mu^*  \rangle ] \\
	&= \conj(\mu^* ) 
	- \conj(\mu_n)  + \E[\bregmanconj(\mu_n, \MAPm)]
	+ \langle \E[\hat \natp_n] \pm \nabla \conj(\mu_n) ;  \mu_n - \mu^* \rangle
	\nonumber\\
	& = \bias{\bregmanconj(\mu^*; \mu_n)}
	+ \frac{n_0}{n+n_0} \langle  \bias{\mu_0 - \mu_*} ; \variance{\E[\hat \natp_n] - \natp_n} \rangle 
	+ \variance{\E[\bregmanconj(\mu_n, \MAPm)]}
\end{align}
With Jensen, we were losing a Bregman term, as per the 3 point property, the covariance is equal to the symmetrized Bregman.
\begin{align}
	\Cov(\hat \natp_n ; \hat \mu_n )  
	=\expect{\bregmanconj(\MAPm ; \mu_n)
	+ \bregmanconj(\mu_n ; \MAPm)}
\end{align}

\paragraph{No Mixed Term.}
In fact, it is possible to have no mixed term using the central point $\tilde \mu_n = \nabla \logpart(\E[\MAPt])$ instead of $\mu_n = \E[\MAPm]$,  exactly as described by \citet[Theorem 0.1]{pfau2013generalized}.
\begin{align}
	\expect{\bregmanconj(\mu^* ; \hat \mu_n)} 
	= \bias{\bregmanconj(\mu^* ; \tilde \mu_n)}
	+ \variance{\expect{\bregmanconj(\tilde \mu_n ; \MAPm)}}
\end{align}
Now the big question, is : what kind of results do I get with such quantities ? In general, it seems that $\tilde \mu_n$ is a very central and interesting quantity to consider.  We show these decompositions in Figure~\ref{fig:experimental_decomposition}.

\begin{figure}[ht]
	\centering
	\includegraphics[width=\textwidth]{{figs/biasvariance/ratio=0.8_n0=10}.pdf}
	\caption{
	\textbf{Left:} training curves and analytic upper bound. 
	\textbf{Center:} bias-mixed-variance decomposition, using the arithmetic mean.
	\textbf{Right:} bias-variance decomposition, using the harmonic mean.
	}
	\label{fig:experimental_decomposition}
\end{figure}

@@

\begin{example}
	We can also get a closed form for the posterior of a centered gaussian variance 
	\begin{align}
		\expect[\pi(\natp|\cD)]{\bregman(\natp ; \natp_*)} = \bregman(\hat \natp_n ; \natp_*) + \frac{2}{n} \frac{\mu_*}{\hat \mu_n} + \log\frac{n}{2} - \psi(1 + \frac{n}{2})
	\end{align}
	using the mean of the \href{https://en.wikipedia.org/wiki/Inverse-gamma_distribution}{inverse Gamma distribution} and the \href{https://en.wikipedia.org/wiki/Gamma_distribution#Logarithmic_expectation_and_variance}{logarithmic expectation} of a Gamma. The using results from the MLE, we get the exact formula
	\begin{align}
		\expect[\cD\sim\natp_*]{\expect[\pi(\natp|\cD)]{\bregman(\natp ; \natp_*)}}
		= \frac{2}{n} + \frac{2}{n(n-2)} + \psi(\half[n]) - \psi(1+ \half[n])
	\end{align}
	and we bound it to 
	\begin{align}
		\inv{2n} + \frac{4}{n(n-2)}
		\leq \expect[\cD\sim\natp_*]{\expect[\pi(\natp|\cD)]{\bregman(\natp ; \natp_*)}}
		\leq 
		\inv{n} + \frac{4}{n(n-2)}
	\end{align}
	which is about twice the MLE bounds~\eqref{eq:MLE_rate}, exactly like as we observed for the gaussian mean.
\end{example}
